Towards Systematic Specification and Verification of Fault-Tolerant Protocols

### 2016-07-25 doc/om-paper/paper.tex
om1 paper updates

Fault-tolerant distributed systems are essential to life-critical systems, like avionics. Consequently, this class of systems demand high-assurance of correct design and implementation. Formal verification can be used to prove correctness.

There is substantial literature on verifying fault-tolerant distributed systems~\cite{}. Some of this work uses mechanical theorem-proving~\cite{}. Two problems with mechanical verification are that it requires specialized expertise and it is labor intensive. In contrast, model-checking is automated but has limited scalability. To address scalability, abstractions and optimizations are employed, both of which lead to problems. Verification abstractions are often ad-hoc and system-specific. For example, to model a process that can broadcast the values $\{0, 1\}$, or no value if when faulty, one might define an enumerated type of broadcast values sent such as

@@

Our contributions combine a number of ideas in the formal verification literature to build scalable, generic formal models of fault-tolerant distributed systems.

@@

We demonstrate the model with an extended case-study, verifying the Oral Messages (1) protocol (OM(1))~\cite{}. While our model is more detailed---it explicitly models channels---and also more general---it allows bounded asynchrony between processes---it is the most scalable model-checking verification of the protocol.

@@

\subsection{Calendar Automata}\label{sec:calendar}\label{sec:calendar}
\emph{Timed automata} are specialized formalisms for verifying real-time systems~\cite{}. Time automata assume the existence of continuously-varying real-valued variables, known as \emph{real-time clocks} in a solver. While successful timed automata model-checkers exist, they are specialized for verifying real-time systems in which the only infinite-valued variables are the real-time clocks. Modern general-purpose model-checkers like \lee{name some} use SMT solvers for general infinite-state system verification that is not limited to real-time clocks.

@@

Define a set of \emph{events} $e_0, e_1, \ldots, e_n \in E$. We do not define events; intuitively, an event is a set of state variables. When an event is \emph{enabled}, the transitions over events are enabled; otherwise, the variables stutter and maintain the same value.

An \emph{event calendar} $\{ (e_0, t_0), (e_1, t_1), \ldots, (e_n, t_n) \}$ is a set of ordered pairs $(e_i, t_i)$ called \emph{calendar events} where $e_i \in E$ is an event and $t_i \in \mathbb{R}$ is a \emph{timeout}, the time at which the event is scheduled. We denote element $(e_1, t_1)$ of an event calendar by $c_i$.

@@

In a distributed system, it is convenient to distinguish global actions and local actions. Global actions are principally interprocess communication, while local actions are those carried out by each process to update its local state and produce new messages to broadcast. While both global and local actions can both be modeled as events in a calendar automata, doing so is generally overkill and complicates the model. Real-time constraints matter at the system-level but rarely within an individual process. From the global perspective, individual processes can update their local state atomically.

@@

Assume processes are indexed from a finite set $Id \subset \mathbb{N}$. A \emph{channel} from process $i$ to $j$ is an ordered pair $(i,j)$. Define a set of messages $Msg$, with a designated element $null \in Msg$ denoting the absence of a message. Given a channel and a timeout, let $send$ return the message sent on the channel at that time:
$$send : Id \times Id \rightarrow \mathbb{R} \rightarrow Msg$$
So $send(i, j, t)$ returns the message that $i$ sends to $j$ at time $t$. Likewise, let
$$recv : Id \times Id \rightarrow \mathbb{R} \rightarrow Msg$$
compute the message received on that channel at that time, so that $recv(i, j, t)$ returns the message received by $j$ from $i$ at time $t$.

In the non-faulty case, we require that messages received were previously sent and not previously received: if $recv(i, j, t) \neq null$, then $\exists t'$ such that $send(i, j, t')$ where $t' < t$, $recv(i, j, t) = send(i, j, t')$, and $\neg\exists t''$ such that $t' < t'' < t$ and $recv(i, j, t'') = recv(i, j, t)$. (We address faults in Section~\ref{sec:kibitzer}.)

Then an event calendar for sending and receiving messages on channels is a set of send or receive events together with their timeouts that yield non-null messages; that is,
$$
\{ (e, t) | \forall i,j: e \neq null \textnormal{ and } (e = send(i, j, t) \textnormal{ or } e = recv(i, j, t))
\}
$$

\lee{technically, atomic time is just a convenience.}
The event of receiving a message initiates a process to update its local state machine and generate additional messages to send. When the process is updating its local state machine, the event calendar is paused. That is, updating an event $recv(i, j, t)$ implicitly includes updating $j$'s state machine.

@@

@@

Not only is modeling faults expensive, but it is generally non-compositional. For example, to model-check the Oral Messages algorithm~\cite{} that provides interactive consistency in the presence of Byzantine faults, Rushby every transition of the General and Lieutenant processes acts on their fault state~\cite{}.

>>>>

Fault-tolerant distributed systems are essential to reliable life-critical systems, like avionics. Consequently, this class of systems demand high-assurance of correct design and implementation. Formal verification can be used to prove correctness.

There is substantial literature on verifying fault-tolerant distributed systems~\cite{}. Most of this work uses interactive theorem-proving~\cite{}. Interactive theorem-proving is useful in the domain because of the complexity of the models; automated methods are difficult to scale. However, two problems with interactive verification are that it requires specialized expertise and it is labor intensive.

In contrast, model-checking is automated but has limited scalability. To address scalability, abstractions and optimizations are employed, both of which lead to problems.

Verification abstractions are often ad-hoc and system-specific. For example, to model a process that can broadcast the values $\{0, 1\}$, or no value when faulty, one might define an enumerated type of broadcast values sent such as

@@

Our contributions combine a number of ideas in the formal verification literature to build scalable, generic formal models of fault-tolerant distributed systems in which verification is more automated than in traditional interactive theorem-proving.

@@

We validate the model with a number of small examples \lee{todo}, as well as an extended case-study, verifying the Oral Messages (1) protocol (OM(1))~\cite{}. While our model is more detailed---it explicitly models channels---and also more general---it allows bounded asynchrony between processes---it is the most scalable model-checking verification of the protocol.

@@

\subsection{Calendar Automata}\label{sec:calendar}\label{sec:calendar}
\emph{Timed automata} are specialized formalisms for verifying real-time systems~\cite{}. Timed automata assume the existence of continuously-varying real-valued variables, known as \emph{real-time clocks} in a solver. Timed automata model-checkers are specialized for verifying real-time systems in which the only infinite-valued variables are the real-time clocks. Modern general-purpose model-checkers like \lee{name some} use SMT solvers for general infinite-state system verification that is not limited to real-time clocks.

@@

Define a set of \emph{events} $e_0, e_1, \ldots, e_n \in E$. For now, we do not define events; intuitively, an event is a set of state variables (shortly, we will associate events with messages sent in a distributed system). When an event is \emph{enabled}, the transitions over events are enabled; otherwise, the variables stutter and maintain the same value.

An \emph{event calendar} $\{ (e_0, t_0), (e_1, t_1), \ldots, (e_n, t_n) \}$ is a set of ordered pairs $(e_i, t_i)$ called \emph{calendar events} where $e_i \in E$ is an event and $t_i \in \mathbb{R}$ is a \emph{timeout}, the time at which the event is scheduled. We denote element $(e_i, t_i)$ of an event calendar by $c_i$.

@@

In a distributed system, it is convenient to distinguish global actions and local actions. Global actions are principally interprocess communication, while local actions are those carried out by each process to update its local state and produce new messages to broadcast. While both global and local actions can both be modeled as events in a calendar automata, doing so is generally overkill and complicates the model. From the global perspective, individual processes can update their local state atomically.

@@

Assume processes are indexed from a finite set $Id \subset \mathbb{N}$. A \emph{channel} from process $i$ to $j$ is an ordered pair $(i,j)$. Define a set of messages $Msg$. Given a channel and a timeout, let $send$ be a relation on messages sent on a channel at a given time:
$$send \subseteq Id \times Id \times \mathbb{R} \times Msg$$
So $send(i, j, t, m)$ holds iff $i$ sends to $j$ message $m$ at time $t$. Likewise, let
$$recv \subseteq Id \times Id \times \mathbb{R} \times Msg$$
be a relation on messages received on a channel at a time, so that $recv(i, j, t, m)$ holds iff the message $m$ received by $j$ from $i$ at time $t$.

In the non-faulty case, we require that messages received were previously sent and not previously received: if $recv(i, j, t, m)$, then $\exists t'$ such that $send(i, j, t', m)$ where $t' < t$, and $\neg\exists t''$ such that $t' < t'' < t$ and $recv(i, j, t'', m) = recv(i, j, t, m)$. (We address faults in Section~\ref{sec:kibitzer}.)

Then an event calendar for sending and receiving messages on channels is the union of the $send$ and $recv$ relations.

\lee{technically, atomic time is just a convenience.}
The event of receiving a message initiates a process to update its local state machine and generate additional messages to send. When the process is updating its local state machine, the event calendar is paused. That is, updating an event $recv(i, j, t, m)$ also includes updating $j$'s state machine.

@@

\lee{todo}

@@

Not only is modeling faults expensive, but it is generally non-compositional. For example, in a classic model-checking model of the fault-tolerant Oral Messages algorithm~\cite{}, every transition of the sending processes updates environmental state variables associated with the fault state.
